---
title: "Gun Control in the U.S.A., Does it Matter?"
author: Darwhin Gomez
format: 
  revealjs:
    center: true
    max-scale: 2.0
    slide-number: true
    toc: false
    theme: simple
    transition: fade
    incremental: false
    beamer: default
    pptx: default
execute:
  echo: false
---

```{r libraries_setup, include=FALSE}
library(tidyverse)
library(readxl)
library(feasts)
library(writexl)
library(maps)
library(scales)
library(usmap)
library(snakecase)
library(gridExtra)
library(ggrepel)
library(rpart)
library(rpart.plot)
library(viridis)
theme_set(theme_minimal())  # Set minimal theme for all ggplots

knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE, message = FALSE,
                      fig.width = 16, fig.height = 9, include = FALSE )

```

```{r datas, include=FALSE, echo=FALSE}
#checking out the sets
gun_mortality <- read.csv("firearm_mortality2022.csv")
str(gun_mortality)
burglary<-read.csv("burglary.csv")

str(burglary)
gun_own_state <- read.csv("gun-ownership-by-state.csv")
gun_laws_states<- read.csv("firearm_laws.csv")
str(gun_own_state)
glimpse(gun_laws_states)
gun_deaths_state<-read.csv("mortality_breakdown.csv")
str(gun_deaths_state)
```

```{r warangling}
# some wrangling
states<- gun_laws_states$STATE
# Build a reference lookup table
state_lookup <- data.frame(
  state_name = state.name,
  state_code = state.abb,
  stringsAsFactors = FALSE
)
# I want to deal with state codes
# Wrangling
df_list <- list(gun_own_state, gun_mortality,gun_laws_states,gun_mortality)

#formatting names
names(gun_deaths_state) <- to_lower_camel_case(names(gun_deaths_state))
names(gun_laws_states) <- to_lower_camel_case(names(gun_laws_states))
names(gun_mortality) <- to_lower_camel_case(names(gun_mortality))
names(gun_own_state) <- to_lower_camel_case(names(gun_own_state))

# function to change state from names to codes
gun_own<-gun_own_state %>%
  mutate(state = str_trim(str_to_title(state))) %>%
  left_join(state_lookup, by = c("state" = "state_name")) %>%
  select(-state) %>%
  rename(state = state_code) 
 
gun_laws<-gun_laws_states %>%
  mutate(state = str_trim(str_to_title(state))) %>%
  left_join(state_lookup, by = c("state" = "state_name")) %>%
  select(-state) %>%
  rename(state = state_code) 
head(gun_laws$state)
gun_laws <- gun_laws %>%
  select(1, state,everything() ) 
gun_laws <- gun_laws %>%
  mutate(lawTotal = as.numeric(lawtotal))

head(gun_laws)
gun_mortality <- gun_mortality |>
  mutate(
    rate = as.numeric(rate),
    deaths = as.numeric(deaths)
  )
my_guns<- gun_laws|>
  select(state,lawtotal)
names(my_guns)


```

```{r maps1}


colonies <- c("DE", "PA", "NJ", "GA", "CT", "MA", "MD",
              "SC", "NH", "VA", "NY", "NC", "RI")

# Example dataset with values for each colony
colonies_data <- data.frame(
  state = colonies,
  value = runif(length(colonies), 10, 100)  # Replace with your actual data
)

map_13 <- plot_usmap(include = colonies, fill = "royalblue") +
  labs(title = " United States in 1776") +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, size = 24)
  )

map_13

# Plot all 50 states in the same color
us_full_map <- plot_usmap(fill = "royalblue") +
  labs(title = "United States 2025") +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, size = 24)
  )

us_full_map


```

```{r maps2, warning=FALSE}
gun_rate_avg <- gun_mortality %>%
  group_by(state) %>%
  summarise(avg_rate = mean(rate, na.rm = TRUE)) %>%
  ungroup()

gun_rate_avg <- gun_rate_avg|>
    filter(state != "District of Columbia")

my_guns <- inner_join(gun_own, my_guns, by = "state")
my_guns<- inner_join(gun_deaths_state,my_guns, by="state")

my_guns<- inner_join(gun_rate_avg,my_guns,by ="state")

us_avg_rate <- plot_usmap(data = gun_rate_avg, values = "avg_rate", regions = "states", labels = TRUE) +
  scale_fill_viridis_c(
    name = "Avg. Gun Death Rate\n(2005–2022)",
    label = scales::comma,
    option = "magma"  # 
  ) +
  labs(
    title = "Average Gun Death Rate by State (2005–2022)",
    caption = "Source: CDC Gun Mortality"
  ) +
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    plot.caption = element_text(size = 10),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  )


us_avg_rate


us_gun_laws_map <- plot_usmap(data = gun_laws, values = "lawTotal", regions = "states", labels=TRUE) +
  scale_fill_viridis_c(
    name = "Total Gun Laws",
    label = scales::comma,
    option = "magma",
    direction = -1
  ) +
  labs(
    title = "Total Gun Laws by State (2019)",
    caption = "Source: https://www.icpsr.umich.edu/web/NACJD/studies/37363/versions/V1/datadocumentation"
  ) + 
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    plot.caption = element_text(size = 10),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  )

us_gun_laws_map
 grid.arrange(
  grobs= list(us_avg_rate,us_gun_laws_map),
  ncol(2),
  top= "Side by Average Death rate/100,000 and total gun control laws"
)

us_gun_own_map <- plot_usmap(data = gun_own, values = "gunOwnershipPercentage" , regions = "states", labels=TRUE) +
  scale_fill_viridis(
    name = "Gun Ownership Percentage",
   label = scales::comma,
    option = "magma",
   
  ) +
  labs(
    title = "State  Gun Ownership Percentage of Poulation",
  ) + 
  theme(
    plot.title = element_text(size = 24, face = "bold"),
    plot.caption = element_text(size = 10),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  )
us_gun_own_map
us_gun_laws_map
 grid.arrange(
  grobs= list(us_avg_rate,us_gun_laws_map),
  ncol(2),
  top= "Side by Average Death rate/100,000 and total gun control laws"
)
 
```

```{r modeling}
#  avg death rates by gun ownership

  deaths_own <- ggplot(data = my_guns, aes(x = gunOwnershipPercentage, y = avg_rate, label= state)) +
  geom_text(color = "royalblue") +
   geom_smooth(method = "lm", color ="darkred", se= TRUE)+
  labs(
    title = " Average Gun Death Rate(2005-2022) by Gun Ownership % by State",
    x = "Gun Ownership (%)",
    y = "AVg Gun Death Rate",
    
  ) + theme(
    plot.title = element_text(size = 18, face = "bold"),
    plot.caption = element_text(size = 10),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  )
deaths_own






deaths_own_lm<-lm(avg_rate~ gunOwnershipPercentage, data=my_guns, method = "lm")


summary(deaths_own_lm)
# death rate by gun laws

deaths_laws<- ggplot(data = my_guns, aes(x = lawtotal, y = avg_rate, label= state)) +
  geom_text(color = "royalblue") +
   geom_smooth(method = "lm", color ="darkred", se= TRUE)+
  labs(
    title = " Average Gun Death Rate(2005-2022) by Total Gun Laws by State",
    x = "State Gun Laws",
    y = "AVg Gun Death Rate",
    
  )+
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    plot.caption = element_text(size = 10),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  )
deaths_laws

gun_law_lm<-lm(avg_rate ~ lawtotal , data = my_guns )


gun_death_ratelm<- lm(avg_rate ~ lawtotal + gunOwnershipPercentage, data = my_guns)

summary(gun_death_ratelm)

own_laws<- ggplot(data = my_guns, aes(x = lawtotal, y = gunOwnershipPercentage, label= state)) +
  geom_text(color = "royalblue") +
   geom_smooth(method = "lm", color ="darkred", se= TRUE)+
  labs(
    title = " Gun Ownership and Gun Laws  by State",
    x = "State Gun Laws",
    y = "State Gun Ownership Percentage",
    
  )+
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    plot.caption = element_text(size = 10),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  )
own_laws

```

## A Nation is Born

::: notes
In 1776, the Thirteen Colonies declared independence from the British Empire. To protect against future threats, the new United States added the Second Amendment to its Constitution:

**"A well regulated Militia, being necessary to the security of a free State, the right of the people to keep and bear Arms shall not be infringed."**
:::

::: panel-tabset
## 1776

```{r slide2a, include=TRUE,width=4, height=7, }

map_13
```

## Today

```{r slide2b, include=TRUE}
us_full_map
```
:::

## Second Amendment in practice today. {style=".tiny"}

::: notes
In the centuries since, the United States has grown, evolved, and now stands as a premier global power. While the Constitution has seen many amendments, the **Second Amendment**—the right to bear arms—remains deeply valued across the nation, though its significance varies from state to state.
:::

```{r slide3,include=TRUE}
us_gun_own_map

```

## State Laws {style=".tiny"}

::: notes
While the Constitution and federal laws continue to uphold the citizens' right to bear arms, it's up to each individual state to set its own laws, regulations, and requirements to legally own a firearm. Some states have way more gun laws on the books than others.
:::

```{r slide4,include=TRUE}
us_gun_laws_map
```

## Gun Deaths Average Rate 2005-2022 {style=".smaller"}

::: notes
It should come as no surprise that guns have a direct link to violence and death. But do states with fewer gun regulations—less laws—actually see more gun-related deaths?
:::

```{r slide5, include=TRUE}
us_avg_rate

```


## Does number of  gun laws influence gun death rate in states?

::: panel-tabset
## Plot

```{r slide7a, include=TRUE}
deaths_laws
```

## linear formula

```{r slide7b, include=TRUE}
summary(gun_law_lm)
```
:::
## Gun Ownership, Death Rate

::: panel-tabset
## Plot

```{r slide6a, include=TRUE}
deaths_own
```

## Linear Formula

```{r slide6b, include=TRUE}
summary(deaths_own_lm)
```
:::


## Combined approach

::: panel-tabset
## Combined model

```{r slide8a, include=TRUE}
summary(gun_death_ratelm)
```
## Gun ownership, Gun laws
```{r slide8b, include=TRUE}
own_laws
```

:::

## Thoughts
::: panel-tabset
## Interpretations

-   The average gun death rate tends to increase as that gun ownership as a percentage of population increases.

-   State with more laws regulating guns experience lower average gun death rate
- We see states with higher gun law counts have smaller a smaller gun ownership percentage.

## Conluding notes

-   Using Ownership to predict average gun death rates accounts for 68% of variance in our combined data set, adding total laws to the model did not improve the models performance.

- By limiting the number of people who own guns we can significantly reduce the average number of gun deaths.

-   Thank You!

